Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016

### 2019-10-08 anzjs_elections.Rmd
added leverage to influence/outlier section

# Combine and standardize
my_df <- bind_rows(
  left_join(eechidna::tpp01,
    standardise_vars(abs2001) %>%
      dplyr::select(-c(UniqueID, Area, ends_with("NS"), Population)),
    by = c("DivisionNm", "StateAb" = "State")
  ) %>% mutate(year = "2001"),
  left_join(eechidna::tpp04,
    standardise_vars(abs2004),
    by = c("DivisionNm")
  ) %>% mutate(year = "2004"),
  left_join(eechidna::tpp07,
    standardise_vars(abs2007),
    by = c("DivisionNm")
  ) %>% mutate(year = "2007"),
  left_join(eechidna::tpp10,
    standardise_vars(abs2010),
    by = c("DivisionNm")
  ) %>% mutate(year = "2010"),
  left_join(eechidna::tpp13,
    standardise_vars(abs2013),
    by = c("DivisionNm")
  ) %>% mutate(year = "2013"),
  left_join(eechidna::tpp16,
    standardise_vars(abs2016) %>%
      dplyr::select(-c(UniqueID, Area, ends_with("NS"), Population)),
    by = c("DivisionNm", "StateAb" = "State")
  ) %>% mutate(year = "2016")
) %>%
  mutate(year = factor(year),
    Band_00_19 = Age00_04 + Age05_14 + Age15_19,
    Band_20_34 = Age20_24 + Age25_34,
    Band_35_54 = Age35_44 + Age45_54,
    Band_55plus = Age55_64 + Age65_74 + Age75_84 + Age85plus) %>%
  select(-c(
    starts_with("Age"), MedianAge, StateAb,
    LNP_Votes, ALP_Votes, ALP_Percent, TotalVotes, Swing,
    InternetUse, InternetAccess, EnglishOnly,
    Other_NonChrist, OtherChrist, Catholic, Anglican,
    Volunteer, EmuneratedElsewhere, Population, UniqueID
  ))

@@

# Order electorates in alphabetical order to match spatial matrix
model_df <- small_df %>%
  arrange(year, DivisionNm) %>%
  dplyr::select(order(colnames(.))) %>%
  rename(
    ManagerAdmin = ManagerAdminClericalSales,
    OtherLanguage = OtherLanguageHome,
    BornSEEuro = Born_SE_Europe,
    BornAsia = Born_Asia,
    BornUK = Born_UK,
    BornMidEast = Born_MidEast,
    OneParentHouse = OneParent_House,
    Pop_00_19 = Band_00_19,
    Pop_20_34 = Band_20_34,
    Pop_35_54 = Band_35_54
  ) %>%
  select(-Band_55plus)
```

@@

```{r cooksd, eval = FALSE}
cooksd_df <- data.frame(
  Residuals = c(glsmod16$actual_residuals, glsmod13$actual_residuals, glsmod10$actual_residuals, glsmod07$actual_residuals, glsmod04$actual_residuals, glsmod01$actual_residuals),
  gls_residuals = c(glsmod16$residuals, glsmod13$residuals, glsmod10$residuals, glsmod07$residuals, glsmod04$residuals, glsmod01$residuals),
  cooksd = c(glsmod16$cooksd, glsmod13$cooksd, glsmod10$cooksd, glsmod07$cooksd, glsmod04$cooksd, glsmod01$cooksd),
  bind_rows(
    glsmod16$my_data %>% left_join(tpp16 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"),
    glsmod13$my_data %>% left_join(tpp13 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"),
    glsmod10$my_data %>% left_join(tpp10 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"),
    glsmod07$my_data %>% left_join(tpp07 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"),
    glsmod04$my_data %>% left_join(tpp04 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"),
    glsmod01$my_data %>% left_join(tpp01 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"))
)
```

@@

Based on the distribution of the Cook's distance values, a Cook's distance greater than $0.1$ is considered to be influential and a potential outlier. The electorate of Sydney (NSW) has a large Cook's distance from 2001 to 2010, due to its diverse population (language, birthplace and religion), high number of defacto relationships, high income, high household mobility and small amount of workers in extractive and transformative jobs. It has remained a strong supporter of the Labor party and the Liberal vote is severely overpredicted by the model, making it an outlier. Nearby in metropolitan NSW, the electorate of Wentworth is found to be an outlier in the 2013 and 2016 elections. Although historically Liberal, its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup --- implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. In the elections since, the model underpredicts Wentworth's Liberal support.

Lingiari, an electorate taking up almost all of the Northern Territory, is an outlier in all but the 2013 election due to its large Indigenous population, young age profile and low rates of property ownership. Fowler (NSW) has a diverse population with a high proportion of migrants, many Buddhists and Muslims, and has strong Labor support, making it influential in 2001, 2004 and 2010. Other electorates with large Cook's distance are Canberra (ACT) and Durack (WA) in 2013, and Solomon (NT) in 2016.

>>>>

# Combine and standardize
my_df <- bind_rows(
  left_join(eechidna::tpp01,
    standardise_vars(abs2001) %>%
      dplyr::select(-c(UniqueID, Area, ends_with("NS"), Population)),
    by = c("DivisionNm", "StateAb" = "State")
  ) %>% mutate(year = "2001"),
  left_join(eechidna::tpp04,
    standardise_vars(abs2004),
    by = c("DivisionNm")
  ) %>% mutate(year = "2004"),
  left_join(eechidna::tpp07,
    standardise_vars(abs2007),
    by = c("DivisionNm")
  ) %>% mutate(year = "2007"),
  left_join(eechidna::tpp10,
    standardise_vars(abs2010),
    by = c("DivisionNm")
  ) %>% mutate(year = "2010"),
  left_join(eechidna::tpp13,
    standardise_vars(abs2013),
    by = c("DivisionNm")
  ) %>% mutate(year = "2013"),
  left_join(eechidna::tpp16,
    standardise_vars(abs2016) %>%
      dplyr::select(-c(UniqueID, Area, ends_with("NS"), Population)),
    by = c("DivisionNm", "StateAb" = "State")
  ) %>% mutate(year = "2016")
) %>%
  mutate(year = factor(year),
    Band_00_19 = Age00_04 + Age05_14 + Age15_19,
    Band_20_34 = Age20_24 + Age25_34,
    Band_35_54 = Age35_44 + Age45_54,
    Band_55plus = Age55_64 + Age65_74 + Age75_84 + Age85plus) %>%
  dplyr::select(-c(
    starts_with("Age"), MedianAge, StateAb,
    LNP_Votes, ALP_Votes, ALP_Percent, TotalVotes, Swing,
    InternetUse, InternetAccess, EnglishOnly,
    Other_NonChrist, OtherChrist, Catholic, Anglican,
    Volunteer, EmuneratedElsewhere, Population, UniqueID
  ))

@@

# Order electorates in alphabetical order to match spatial matrix
model_df <- small_df %>%
  arrange(year, DivisionNm) %>%
  dplyr::select(order(colnames(.))) %>%
  rename(
    ManagerAdmin = ManagerAdminClericalSales,
    OtherLanguage = OtherLanguageHome,
    BornSEEuro = Born_SE_Europe,
    BornAsia = Born_Asia,
    BornUK = Born_UK,
    BornMidEast = Born_MidEast,
    OneParentHouse = OneParent_House,
    Pop_00_19 = Band_00_19,
    Pop_20_34 = Band_20_34,
    Pop_35_54 = Band_35_54
  ) %>%
  dplyr::select(-Band_55plus)
```

@@

```{r cooksd, eval = FALSE}
cooksd_df <- data.frame(
  Residuals = c(glsmod16$actual_residuals, glsmod13$actual_residuals, glsmod10$actual_residuals, glsmod07$actual_residuals, glsmod04$actual_residuals, glsmod01$actual_residuals),
  gls_residuals = c(glsmod16$residuals, glsmod13$residuals, glsmod10$residuals, glsmod07$residuals, glsmod04$residuals, glsmod01$residuals),
  cooksd = c(glsmod16$cooksd, glsmod13$cooksd, glsmod10$cooksd, glsmod07$cooksd, glsmod04$cooksd, glsmod01$cooksd),
  Hat = c(glsmod16$hatvalues, glsmod13$hatvalues, glsmod10$hatvalues, glsmod07$hatvalues, glsmod04$hatvalues, glsmod01$hatvalues),
  bind_rows(
    glsmod16$my_data %>% left_join(tpp16 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"),
    glsmod13$my_data %>% left_join(tpp13 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"),
    glsmod10$my_data %>% left_join(tpp10 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"),
    glsmod07$my_data %>% left_join(tpp07 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"),
    glsmod04$my_data %>% left_join(tpp04 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"),
    glsmod01$my_data %>% left_join(tpp01 %>% dplyr::select(DivisionNm, UniqueID, StateAb), by = "DivisionNm"))
)
```

```{r plothatandcooksd, eval = FALSE}
cooksd_df %>%
  ggplot(aes(x = cooksd, y = Hat)) +
  geom_point(shape = 1) +
  #geom_text(aes(label = DivisionNm), size = 3) +
  facet_wrap(~year, nrow = 2) +
  # coord_fixed(ratio = .5) +
  labs(x = "Cook's Distance", y = "Hat values") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), axis.text.y = element_text(size = 8)) +
  theme_bw()
```

@@

Based on the distribution of the Cook's distance values and the distribution of hat values, a Cook's distance greater than $0.1$ is considered to be influential, and a hat value greater than $0.5$ is considered to have high leverage. Electorates fitting these criteria are flagged and investigated to examine the characteristics driving these values.

The electorate of Sydney (NSW) has a large Cook's distance and high leverage from 2001 to 2007, due to its diverse population (languages, birthplace and religion), high density of young adults (20 to 34 years old), high number of defacto relationships, high income, high household mobility and small amount of workers in extractive and transformative jobs. It has remained a strong supporter of the Labor party and the extent of this support is underpredicted by the model, making it an outlier. Nearby in metropolitan NSW, the electorate of Wentworth is found to be an outlier in the 2013 and 2016 elections. Although historically Liberal, its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup --- implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. In the elections since, the model underpredicts Wentworth's Liberal support.

Lingiari, an electorate taking up almost all of the Northern Territory, has consistently high leverage (all years) and is an outlier in all but the 2013 election due to its large Indigenous population, low rates of property ownership and few workers in management or administrative jobs. Fowler (NSW) has a diverse population with a high proportion of migrants, many Buddhists and Muslims, as well as a high proportion of single parent households. These characteristics explain its high leverage in 2001, 2004, 2010 and 2013, and its strong Labor support makes it influential in 2001, 2004 and 2010. Other electorates with large Cook's distance are Canberra (ACT) and Durack (WA) in 2013, and Solomon (NT) in 2016. All of the electorates examined are not unduly influential in the model and therefore no action is required.
